Dimension Splitting and a Long Time-Step Multi-Dimensional Scheme for Atmospheric Transport

### 2017-06-07 revision/reply_to_reviewer.tex

\documentclass[12pt]{article}
\usepackage[margin=2cm]{geometry}
\usepackage{url,times}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[authoryear]{natbib}
\usepackage{todonotes}
\usepackage[normalem]{ulem}
\usepackage{color}

\begin{document}

\title{Responses to reviewer comments on article QJ-17-0017\\
Dimension Splitting and a Long Time-Step Multi-Dimensional Scheme for Atmospheric Transport}
\author{Yumeng Chen \and Hilary Weller \and Stephen Pring \and and James Shaw}
\maketitle

@@

\item [Comment 1]: This is the most minor, but there is a study for semi-Lagrangian dimensionally split transport on the cubed-sphere that needs to be considered and discussed: \cite{GNQ14}. Additionally, \cite{KNK15} was cited as a dimensionally split scheme, but it is not. The reconstruction is performed in dimensional sweeps, but the scheme itself is multi-dimensional.

\item [Response]: 
Thank-you for pointing out the \cite{GNQ14} paper. See response to pdf annotations below.

\item [Comment 2]:
It is stated that results in this 2-D framework will generalize well to the cubed-sphere, but I believe that in order to truly make that statement, one would have to implement a given scheme on both meshes and then compare. My experience on the cubed-sphere mesh is that it.

\item [Response]:
Agreed. Our test cases on a plane will highlight errors due to non-orthogonality and due to discontinuities at cube edges. Some sentences have been added to the introduction:
\begin{quote}
We assume that a large fraction of the numerical errors associated with terrain following co-ordinates and the cubed-sphere grid are due to the mesh non-orthogonality and distortion. Other sources of numerical errors, such as those associated with the representation of spherical geometry, are not addressed here.
\end{quote}
and
\begin{quote}
These test cases will mimic the problems encountered near cubed-sphere edges. 
\end{quote}
A similar caveat has been inserted in the second paragraph of the abstract, the first paragraph of section 3 and in the third paragraph of the conclusions.

@@

\item [Comment 3]: Most importantly, to say the newly proposed dimensionally split scheme is accurate compared to multi-dimensional schemes, I believe a better comparison point is needed. Users select the scheme that gives the best results with the fewest resources. I believe the authors need to include a time-explicit semi-discrete (method of lines) multi-dimensional scheme with cubic interpolants using smaller time steps to serve as a comparison point for this scheme. I know the focus is on large time steps, but it's necessary to consider if one can, in fact, get more accuracy for a given amount of work by simply using the status quo with smaller time steps.

\item [Response]:
An important result from this paper is that the dimensionally split scheme loses very little accuracy when grid distortions are introduced. This result does not rely on the accuracy of the multi-dimensional scheme. Text has been added throughout the paper to clarify this point. The text in blue has been added:
\begin{itemize}
\item In the abstract:
\begin{quote}
Mesh distortions are expected to accentuate the errors associated with dimension splitting, however, \textcolor{blue}{the accuracy of} the dimensionally split scheme \sout{is very accurate on orthogonal meshes and accuracy} decreases only a little in the presence of mesh distortions.
\end{quote}
\item In the introduction:
\begin{quote}
we show the effect of dimension splitting errors using a FFSL dimensionally split scheme on a number of test-cases which use distorted \textcolor{blue}{and undistorted} meshes.
\end{quote}
\item In the conclusions
\begin{quote}
The dimensionally split scheme is most accurate on orthogonal meshes but only loses a little accuracy on highly distorted meshes, despite a first-order departure point calculation.
\end{quote}
\end{itemize}

@@

We also now compare the dimensionally split scheme with the multi-dimensional scheme when the multi-dimensional scheme is run with explicit and implicit time-stepping. New results are described in section 3. The errors of the multi-dimensional scheme (which is now described in full by \cite{SWMD17}) are insensitive to the choice of time-stepping type (implicit or explicit) for Courant numbers smaller than two. These comparisons have also revealed that the implicit solve is a small part of the total cost of the advection. This is now discussed in section 3.4:

\begin{quote}
We also measured the wall-clock computation cost, comparing the implicit and explicit versions of the multi-dimensional scheme running the deformational test case in serial on a non-orthogonal mesh with a maximum Courant number of 1. At the resolution of $120\times 60$, the model using implicit time-stepping is 5\% 
more expensive that the explicit model. This drops to 2.5\%
at the resolution of $960\times 480$. Despite the implicit scheme having four times as many multiplies as calculated for table 3, the difference in wall-clock time is negligible. This demonstrates both the danger of only looking at the number of multiplies and the danger of using results of timing code. The linear equation solvers on OpenFOAM have undergone years of optimisation whereas the stencils for calculating the cubic fit have not been optimised and entail unpredictable memory access patterns and likely frequent cache misses. 
\end{quote}

We have added some description of why we think it suitable to compare the dimensionally split and multi-dimensional schemes at the beginning of section 2:
\begin{quote}
The dimensionally split scheme is third-order accurate in one dimension but the COSMIC splitting reduces the temporal order to two. The multi-dimensional scheme uses cubic interpolations but uses cell centroids as Gauss points so the order of accuracy is limited to two. When explicit time-stepping is used, the cost of both schemes is similar (see section 3.4). These schemes are therefore considered suitable for comparison. 
\end{quote}

@@

Several additions have been made as we are now using implicit and explicit time-stepping for the multi-dimensional scheme:

\begin{itemize}
\item At the beginning of section 2:
\begin{quote}
an explicit, second-order Runge-Kutta time-stepping scheme and with
\end{quote}

\item
Section 2.3.1 has been split into two, the first to describe the implicit and explicit time-stepping and the second to describe the matrix solvers and tolerances.

\item Added to section 3.1:
\begin{quote}
The results in figure 3 use implicit time-stepping but the results using explicit time-stepping are similar (not shown). 
\end{quote}
and
\begin{quote}
The explicit multi-dimensional scheme is not stable for this time-step.
\end{quote}
and
\begin{quote}
Results of the explicit multi-dimensional scheme are also shown in figure 5 for the shorter time-step on the non-orthogonal mesh. The explicit RK2 scheme is more accurate than the implicit (Crank-Nicolson) scheme with the same order of convergence but does not come close to the accuracy of the dimensionally split scheme.
\end{quote}
and
\begin{quote}
The explicit version of the multi-dimensional scheme has errors that decrease as time-step increases due to the spatial discretisation errors that are being introduced at every time-step. 
\end{quote}

@@

\item 
Added to section 3.2:
\begin{quote}
The multi-dimensional scheme uses implicit time-stepping.
\end{quote}

\item 
Added to section 3.3:
\begin{quote}
(with explicit time-stepping although the results using implicit time-stepping look identical)
\end{quote}
and
\begin{quote}
The explicit multi-dimensional scheme is stable at this Courant number and errors are almost identical to those of the implicit scheme.
\end{quote}
and
\begin{quote}
Again this shows that the explicit time-stepping gives almost identical errors to the implicit time-stepping for time-steps at which it is stable.
\end{quote}

Added to the conclusions:
\begin{quote}
Computational cost is estimated by the number of multiplies.\\
...\\
However in practice we found that the implicit solution contributed only 2.5\%-5\% of the total wall-clock computation time.
\end{quote}
\end{itemize}
\end{description}

@@

\begin{enumerate}
\item [Page 1, Line 41]: Might want to say ``distorted, non-orthogonal meshes'' to make it's clear that you are indeed directly testing non-orthogonality.

\item [Response]: Changed as suggested.

\item [Page 2, Line 4]: quasi-uniform resolution doesn't help better exploit modern architectures; they're largely orthogonal concepts. Quasi-uniform resolution just improves the global time step restriction.

\item [Response]: Have changed the beginning of the introduction to:
\begin{quote}
Many traditional weather and climate models use latitude-longitude meshes, but new models are being developed on quasi-uniform meshes in order to better exploit modern computers \cite[eg][]{WTC12,LUJ+14,ST12,SG11,KNK15}. Latitude-longitude meshes need measures to enable long time-steps near the poles such as polar filtering, semi-implicit and semi-Lagrangian methods \cite[]{DCM+05}. These methods necessarily have large domains of dependence near the poles which leads to poor parallel scaling at high resolution. Quasi-uniform meshes avoid parallel scaling bottlenecks near the poles but lead to non-orthogonal meshes (eg the cubed-sphere) or different structures (eg icosahedra).
\end{quote}

@@

\item [Page 2, Line 45]: These guys \cite[]{PL07} appear to be using a sort of predictor-corrector approach where one predicts in the $x$-direction to $\Delta t/2$, then corrects with $y$ over the time-step. Then they swap the dimensions and average the final result. Their errors are certainly more uniform over time, but they are larger (using $90\times 90$ DOFs per panel) than \cite{GNQ14} (who used $80\times 80$ DOFs per panel).

\item [Response]: See below

\item [Page 2, Line 46]: The scheme \cite[]{KNK15} in this paper isn't really dimensionally split, only the *reconstruction* is. \cite{KNK15} use dimensionally split sweeps to create a genuinely multi-dimensional interpolant (minus a few of the cross terms), and then they evolve in a genuinely multi-dimensional way. This is an important distinction because (1) they do not incur time error in the same way as truly split schemes and (2) they include a diagonal sweep for some of the cross terms, whereas truly split schemes generally do not.

\item [Response]: These sentences have been changed to:
\begin{quote}
Dimensionally split schemes have been found to give good accuracy on non-orthogonal meshes such as the cubed-sphere if special treatment is applied over cube edges to improve accuracy: for example, \cite{PL07} use the average of two one-sided schemes at cube edges. \cite{KNK15} use a multi-dimensional transport scheme with dimensionally split reconstructions. They create ghost cells outside each cube panel boundary and achieve accuracy between second and fourth order. \cite{GNQ14} report error growth around cube edges using Strang splitting. 
\end{quote}

@@

\item [Page 3, Line 45]: What do you mean by maximally? Do you mean ``highly'', or is something being optimized here?

\item [Response]: Replaced by:
\begin{quote}
without smoothing of terrain following layers
\end{quote}
Also made a modification in the conclusions:
\begin{quote}
using a basic terrain following mesh in order to maximise the distortions \textcolor{blue}{due to terrain following coordinates}.
\end{quote}

\item [Page 5, Line 1]: Thanks for mentioning this. Do you think using quadrature in time would be worthwhile? It'd have to be separate over each individual cell because it requires continuity, so it'd be pretty expensive

\item [Response]: We have added:
\begin{quote}
The departure point calculation could be improved by using a second-order predictor-corrector departure point calculation \cite[eg][]{MDW+10}.
\end{quote}

@@

\item [Page 10, Line 55]: Have you thought about using a more accurate implicit time stepping than CN? It would probably improve the phase errors.

\item [Response]: Good idea. Simpler than improving the spatial accuracy of the multi-dimensional scheme. However improving the spatial accuracy is a higher priority since for Courant numbers less than one, the multi-dimensional scheme errors are insensitive to time-step. No change made to manuscript.

\item [Page 13, Line 1]: This is really a comparative statement, and I think the baseline is fairly poor.

\item [Response]: We agree that the baseline could be better. However the two schemes do have similar complexity and cost. The final sentence has been changed to:
\begin{quote}
The multi-dimensional scheme is insensitive to the orthogonality, converges with second-order \textcolor{blue}{as expected} and suffers from phase errors at large Courant numbers \textcolor{blue}{due to the use of Crank-Nicolson implicit time-stepping}.
\end{quote}
